---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hmcSandbox

<!-- badges: start -->
<!-- badges: end -->

The hmcSandbox provides a customizable R-implementation of the [NUTS-sampler](https://arxiv.org/abs/1111.4246) which allows the user to extract sampler diagnostics from each algorithm step. 

## Installation

The hmcSandbox is built using the Bridgestan and Cmdstan frameworks, which both needs to be installed before the package can be used. 

Installation links:

* Bridgestan: https://github.com/roualdes/bridgestan

* Cmdstan: https://mc-stan.org/users/interfaces/cmdstan

The development version of hmcSandbox can then be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("JTorgander/hmc-sandbox")
```

## Demonstration 

### Loading models

In order to use a given stan-model within the hmc-sandbox, the corresponding stan file `model_name.stan` needs to be placed in a named folder `model_name`. This folder should also contain the input data in a JSON-file named `model_name.data.json`. This folder should in turn be placed in the folder `test_models` in the Bridgestan directory.

To illustrate this we will use the Bridgestan test model `regression`, defined as follows:

```{r, eval=FALSE}
"data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
transformed parameters {
  vector[N] mu = alpha + beta * x;
}
model {
  alpha ~ normal(0, 5);
  beta ~ normal(0, 3);
  sigma ~ cauchy(0, 1.5);
  y ~ normal(mu, sigma);
}
generated quantities {
  real x_gen = normal_rng(0, 2);
  real y_gen = normal_rng(alpha + beta * x_gen, sigma);
}"
```
For this model, the Bridgestan test_models folder contains a folder `regression`, which in turn contains one model file `regression.stan` and one data file `regression.data.json`.

Before the stan model can be used within Bridgestan, the .stan file needs to be compiled in to compiled shared object `.so` file. For mac users the model can be compiled with the following function.

```{r}
library(hmcSandbox)
BS_PATH <- "~/Documents/bridgestan"
model_name <- "regression"
compile_bs(model_name=model_name, bridgestan_path=BS_PATH)
```

Here the argument "bridgestan_path" should be set to the Bridgestan install directory. Other users are referred to the Bridgestan [documentation]( https://github.com/roualdes/bridgestan). Note that the model needs to be recompiled whenever the `.stan` file is modified. The compiled model can now be loaded using the following function. 

```{r}
bs_model <- get_model(model_name, model_seed = 1234, data=TRUE, sampler_type = "bridgestan", bridgestan_path=BS_PATH)
```
Setting `sampler_type = "bridgestan"` here indicates that the implementation based on bridgestan should be used and setting `data=TRUE` here indicates that the data.json file featured in the model folder should be used when loading the model.

### Sampling

Having loaded the model object, we can now sample from the loaded model using the following code:

```{r bs_sampling}
bs_fit <- NUTS(model = bs_model, n_samples = 4000, warm_up = 2000)
```
This will yield 4000 samples using a burn-in/initialization period of 2000 samples. For reference purposes, the model object also contains the original cmdstan model. This model can in turn be sampled from as before using the `sample` method:

```{r stan_sampling}
stan_fit <- bs_model$model$cmdstan_model$sample(data = bs_model$data, chains = 1, iter_sampling = 4000)
```
### Inspecting samples

We can now inspect and compare the samples generated from our both samplers. The output format of the samplers should work with the [bayesplot](http://mc-stan.org/bayesplot/) library

```{r plotting, message=FALSE}
library(bayesplot)
library(ggplot2)
library(gridExtra)
bs_posterior <- as.array(bs_fit$samples)
stan_posterior <- as.array(stan_fit$draws())

bs_hist <- mcmc_hist(bs_posterior) + labs(title = "Bridgestan")
stan_hist <- mcmc_hist(stan_posterior, pars = c("alpha", "beta", "sigma")) + labs(title = "Stan")
grid.arrange(bs_hist, stan_hist)

```

### Extracting sample trajectories

Given a sampling run, the trajectories of the leapfrog iterations can be extracted as follows. 

```{r}
trajectories <- get_trajectory_stats(bs_fit, n_samples=1)
knitr::kable(trajectories)
```
`
For each sample, parameter and leapfrog algorithm update the following information is currently extracted:

* `subtree`: which binary subtree of the NUTS algorithm the leapfrog step belongs to
* `tree_dir` indicating if the current subtree is grown in a positive or negative direction
* `g1` `g2`: first and second gradient update 
* `p1`, `p2`: first and second momentum updates: 
* `h1_diag`, `h2_diag`: diagonal of first and second Hessian: 
* `h1_param1_param2`, `h2_param1_param2`: lower triangular Hessian components: 
* `g1_l2`, `g2_l2`: $L^2$-norm of the first and second gradient: 
